Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics

### 2013-11-15 paper/paper2.tex
a few paper edits plus added new figures for the MVN likelihood

For example, consider two Markov state models sharing the same transition matrix, $T$. In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states. If an observed trajectory does go through the first models’ regions of high likelihood, it is appropriately termed a more likely model given the data.

However, models' long timescale behavior -- the rates and fluxes between metastable basins -- are independent of the choice of the emission distributions. The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state. If such differences exist and are sufficiently large to warrant attention, than the states are too large.

@@

@@

Instead, we explore both cross validation and Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} for choosing the number of states in a Markov state model. For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluating the likelihood on the left-out portion in an attempt to directly measure the generalization error. The BIC on the other hand involves augmenting the likelihood with a penalty on the number of parameters, and is
an asymptotic approximation Bayesian evidence.

\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + k \cdot \ln N
\end{equation}

@@

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.

The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and furthermore, the likelihood remained almost exactly constant, which gives us some confidence that the volumes do not need to be perfect in order to have a good estimate of the likelihood. \hl{I can make a figure, and we can put it in the SI}.

\begin{figure}[h]
\centering
\includegraphics[width=3.3in]{figs/mull_cross.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states. There are many methods that can be used to test the transferability of an MSM. We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations. These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily. \hl{See SI section: ``AIC vs BIC" for a discussion of the various methods in detail.}}
\label{fig:mullerlike}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=2in]{figs/ww_aic_vs_time.png}
\caption{Implied timescales and AIC scores for a series of Markov state models for the Fip35 WW domain. The panels differ by the dimensionality of the tICA projection. All models were built using a Markovian lag time of 50 ns. Qualitatively, the optimal model according to the AIC, corresponds roughly to where the timescales begin to flatten out.}
\label{fig:ww}
\end{figure}

@@

>>>>

For example, consider two Markov state models sharing the same transition matrix, $T$. In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states. If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.

This emission distribution can be defined in a number of ways. The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics. When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume. This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}. For this reason, we can also choose other emission distributions that may be more tractable. In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings. The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering. If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.

@@

\begin{equation}
\label{eq:like_mvn}
\begin{split}
P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \sigma_{s(x_{i})}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \sigma_{s(x_{i})}^2}\right)
\end{split}
\end{equation} where $\mu_{s(x_i)}$ is the cluster center for the state that $x_i$ is assigned to. For simplicity, we follow \citet{Pelleg2000Xmeans} for estimating the intrastate variances, and simply use a maximum likelihood estimator so that the likelihood becomes:
\begin{equation}
\label{eq:like_mvn}
\begin{split}
P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
\end{split}
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
\end{equation} where $k$ is the number of states.

@@

Instead, we explore cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model. For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluating the likelihood on the left-out portion in an attempt to directly measure the generalization error. The BIC on the other hand involves augmenting the likelihood with a penalty on the number of parameters, and is
an asymptotic approximation Bayesian evidence.

\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + k \cdot \ln N
\end{equation}
\begin{equation}
\label{eq:eic}
\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot k
\end{equation}

@@

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the first $10^6$ steps using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.

The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and furthermore, the likelihood remained almost exactly constant, which gives us some confidence that the volumes do not need to be perfect in order to have a good estimate of the likelihood.

For cross-validation, we used the final $10^6$ frames as the test data.

\begin{figure}[h]
\centering
\includegraphics[width=2in]{figs/mull_cross_vert.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states. There are many methods that can be used to test the transferability of an MSM. We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations. These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily. \hl{See SI section: ``AIC vs BIC" for a discussion of the various methods in detail.}}
\label{fig:mullerlike}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=3in]{figs/kmeans_vor_like.png}
\caption{MSMs were built using K-Means clustering on the M\"uller potential simulations. Since K-Means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model. Interestingly, the likelihood decreases as you introduce additional states. This behavior is due to the state-to-state likelihood decreased rapidly when adding additional states, which may be a result of the large number of states in the most stable well.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs/ww_logLs.png}
\caption{For models built using tICA, we were able to compute volumes as the dimensionality was low (A). These likelihoods were qualitatively similar to the MVN likelihoods (B). Because the MVN likelihood is easy to compute, we can use it to calculate the likelihood of a model clustered using the RMSD distance metric (C). Our likelihood indicates we should construct a model with 5000 states in the RMSD metric, whereas previous studies used a model with 26,000 states \cite{Lane2011Markov}.}
\label{fig:ww}
\end{figure}

@@

% old stuff about using uniform distributions
However, models' long timescale behavior -- the rates and fluxes between metastable basins -- are independent of the choice of the emission distributions. The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state. If such differences exist and are sufficiently large to warrant attention, than the states are too large.
